---
title: Exploring relationship between transmission and other variables for MPG in
  mtcars dataset
author: "Anatoliy Medvedchuk"
date: "25.07.2015"
output:
  pdf_document:
    fig_height: 2.7
    fig_width: 6
  html_document: default
---

#Overview

This report contains discovering the relationship between a set of predictor variables and miles per gallon (MPG) for `mtcars` dataset using linear regression models. The two basic question of interest are:

1. Is an automatic or manual transmission better for MPG?
2. Quantify the MPG difference between automatic and manual transmissions

#Exploratory analysis

`mtcars` dataset structure is as follows:
```{r echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)
library(datasets)
data(mtcars)

head(mtcars)

```

```{r echo=FALSE, warning=FALSE, message=FALSE}
g <- ggplot(data = mtcars, aes(x = am, y = mpg)) +
    geom_boxplot(aes(fill = factor(am))) +
    labs(title = "MPG for manual and automatic transmission")
g
```

Additionl exploratory plots you can find in appendix.

#Regression analysis

Lets try to fit simplest model where `mpg ~ am` and print coefficients (according to `?mtcars` transmission type variable is `am` with meaning: 0 - automatic, 1 - mechanic):

```{r echo=FALSE, warning=FALSE, message=FALSE}
fit <- lm(mpg~am, mtcars)

#summary(lm(mpg~factor(am), mtcars))
#summary(lm(mpg~factor(am)+hp+wt, mtcars))

summary(fit)$coefficients
```
According to P-value coefficients are statistically signifficant so we should reject null hypotesis (H0: there is no differences between MPG in mechanic and automatic transmission). `Intercept` = `r summary(fit)$coefficients[1,1]` and `am` = `r summary(fit)$coefficients[2,1]`. It means expected value of MPG for mechanic is on `r summary(fit)$coefficients[2,1]` greater than MPG for automatic transmission.

But accourding to r.squared value (`r summary(fit)$r.squared`) only about 36% of variation is explained by relation mpg~am. So there are some other dependencies.

Let's fit additional model with different number of predictors. To compare each next one model to previous one Analysis ov variance is used (ANOVA):
```{r echo=FALSE, warning=FALSE, message=FALSE}

#plot(mtcars$am, mtcars$mpg)
#plot(mtcars$hp, mtcars$mpg)
#plot(mtcars$wt, mtcars$mpg)

#fit0 <- lm(mpg ~ hp, data = mtcars)

fit1 <- lm(mpg ~ am, data = mtcars)
fit11 <- update(fit1, mpg ~ hp)
fit2 <- update(fit1, mpg ~ am + hp)
fit3 <- update(fit1, mpg ~ am + hp + wt)
fit31 <- update(fit1, mpg ~ hp + wt)
fit4 <- update(fit1, mpg ~ am + hp + wt + cyl + disp + drat + qsec + vs + gear + carb)

anova(fit1, fit2, fit3, fit4)

# t<-anova(fit1, fit2, fit3, fit4, fit5)

```

From the anova table we can see that residual square sum (RSS) becomes smaller when new variables intoduced as predictors. But for model #4 (including all variables) P-value is too large, so this model can not be statistically significant. So model #3 (mpg ~ am + hp + wt) is most useful model for prediction MPG based on current analysis.
R squared for third model is `r summary(fit3)$r.squared`, `sigma` = `r summary(fit3)$sigma`, coeficients:

```{r echo=FALSE, warning=FALSE, message=FALSE}
summary(fit3)$coefficients
```

#Summary
Answers for questions:

1. Manual transmission is better for MPG than automatic according to `mtcars` data sample.
2. Difference between expected values of MPG for manual and automatic transmission is `r summary(fit)$coefficients[2,1]`

Based on comparing different models with anova method, model `mpg ~ am + hp + wt` was defined as most useful for explaining the data.

#Appendix

## Exploratory plots

```{r echo=FALSE, warning=FALSE, message=FALSE}
g <- ggplot(data = mtcars, aes(x = am, y = mpg)) +
    geom_point(aes(color = factor(am)), size=5) +
    geom_smooth(method = "lm") +
    labs(title = "Regression for manual and automatic transmission")
g

g <- ggplot(data = mtcars, aes(x = hp, y = mpg)) +
    geom_point(aes(color = factor(am)), size=5) +
    geom_smooth(data = mtcars[mtcars$am==1,], method = "lm") +
    geom_smooth(data = mtcars[mtcars$am==0,], method = "lm") +
    labs(title = "Regression for hp as predictor in context of transmission type")
g

g <- ggplot(data = mtcars, aes(x = wt, y = mpg)) +
    geom_point(aes(color = factor(am)), size=5) +
#     geom_smooth(method = "lm") +
    geom_smooth(data = mtcars[mtcars$am==1,], method = "lm") +
    geom_smooth(data = mtcars[mtcars$am==0,], method = "lm") +
    labs(title = "Regression for wt as predictor in context of transmission type")
g
```

## Residual plots

```{r echo=FALSE, warning=FALSE, message=FALSE}

g <- ggplot(data = mtcars, aes(x = am, y = resid(fit))) +
    geom_point(size=5, colour="blue", alpha = 0.4) +
    labs(title = "Residual plot for lm(mpg ~ am)")
g

g <- ggplot(data = mtcars, aes(x = hp, y = resid(fit3))) +
    geom_point(size=5, colour="blue", alpha = 0.4) +
    labs(title = "Residual plot for hp from lm(mpg ~ am + hp + wt)")
g

g <- ggplot(data = mtcars, aes(x = wt, y = resid(fit3))) +
    geom_point(size=5, colour="blue", alpha = 0.4) +
    labs(title = "Residual plot for wt from lm(mpg ~ am + hp + wt)")
g
```

